---
title: Ant data Generalized Linear Model - Frequentist
author: Brett Melbourne
date: 11 Oct 2018 (updated 21 Oct 2021)
output:
    github_document:
        pandoc_args: --webtex
---

This is the first in a series of scripts to analyze the ant data described in Ellison (2004). This script includes plotting variations (ggplot & base) and frequentist inference from the GLM. Future scripts will consider a Bayesian approach to the ant analysis, as well as multilevel models to fully account for the design structure and data variance.

This script can be rendered to a reproducible report.\
`rmarkdown::render("09_5_ants_frequentist.Rmd")`, or *Ctrl+Shift+K* in RStudio.

```{r results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(scales) #for ggplot color palette in function hue_pal
```

## EDA

Read in the ants data:

```{r}
ant <- read.csv("data/ants.csv")
head(ant)
```

For now we will focus on the relationship of species richness with latitude
and habitat. Later we will consider the other variables.

Convert habitat to a factor

```{r}
ant$habitat <- factor(ant$habitat)
```

I will illustrate various approaches to plotting the data in both base graphics and ggplot. First, a basic plot, in the plot.default style of base plot. There are several things to note. 1) We need to extract the columns from the data frame using the `$` operator. 2) We use the graphical parameter `col` to map habitat type to point color. 3) We can make the legend automatically plot the correct
colors for each level of habitat by using the functions `levels` and `nlevels`.
```{r eval=FALSE}
?plot.default #more information here
```
```{r}
plot(ant$latitude, ant$richness, col=ant$habitat)
legend("topright", levels(ant$habitat), col=1:nlevels(ant$habitat), pch=1)
```

The base plot defaults are open circles as points, which is good for visualizing over-plotting, and black and red for the first two colors, which provides reasonable contrast.

We can make a similar plot in the plot.formula style. The two things that change: 1) We use a formula to specify the y and x variables, using standard model formulae notation (`?formula`). 2) The data frame is given in a separate argument. The code for the `plot()` function is now cleaner because we don't need to repeatedly specify the data frame (but the `legend()` function is unchanged).

```{r eval=FALSE}
?plot.formula #more information here
```
```{r}
plot(richness ~ latitude, col=habitat, data=ant)
legend("topright", levels(ant$habitat), col=1:nlevels(ant$habitat), pch=1)
```

The resulting plot is the same as the previous one except with nicer default axis labels.

Another approach is to use the `with()` function to specify the data frame, which has the benefit of working equally with the `legend()` function. This will produce an identical plot to the previous code. 

```{r eval = FALSE}
with(ant, plot(latitude, richness, col=habitat))
with(ant, legend("topright", levels(habitat), col=1:nlevels(habitat), pch=1))
```

As a further exploration of some base plot options, we'll reproduce the original Figure 12a of Ellison, which was in black and white. I will explicitly give all of the variable mappings to highlight a grammar of graphics style. This time we map habitat to plotting character (symbol) using the graphics option `pch` but effectively we are mapping habitat to color (2 colors, black and white), an effective choice because it provides a color scale with maximum contrast for the 2-level factor. To control the plotting character we use the extract function `[]`, which uses the numerical levels of the habitat factor (see `?"["`). We've added nicer axis labels.

```{r}
with(data=ant,
{
    plot(x=latitude, y=richness, pch=c(1,16)[habitat],
         ylab="Ant species richness", xlab="Degrees north latitude")
    legend("topright", levels(habitat), pch=c(1,16))
}
)
```

Now, let's take a look at `ggplot`. The most common, abbreviated, style would be

```{r}
ggplot(ant, aes(latitude, richness, col=habitat)) + 
    geom_point()
```

Clearly this is an attractive and effective default plot. Some of the choices compared to the base plot defaults are the use of a grid to aid judging distance on a common scale, a color scale with good contrast between categories, filled points and a grey background to help the points pop, and an automatic legend placed outside of the plotting area.

Since we are focusing on the grammar of graphics logic and particularly the mapping of variables to visual channels, from here we'll use a slightly more verbose style to emphasize our thoughtful choices. The following code would produce a plot identical to the plot made by the terser code above.

```{r eval=FALSE}
ggplot(data=ant, mapping=aes(x=latitude, y=richness, col=habitat)) + 
    geom_point()
```

A nice feature of `ggplot` is the use of themes. Here is the same plot with the black and white theme.

```{r}
ggplot(data=ant, mapping=aes(x=latitude, y=richness, col=habitat)) + 
    geom_point() +
    theme_bw()
```


You can use a grammar of graphics workflow with base plot functions too.

One of the most recognizable features of ggplot is its default color scale. Here's how to use that color scale with base plot. The color scale is derived from certain settings in the `hue_pal()` function in the `scales` package. We'll set up a new function called `ggc` ("ggplot colors") that, when called with `ggc(n)`, will return a color palette with `n` colors.

```{r}
ggc <- hue_pal(h = c(0, 360) + 15, c = 100, l = 65, h.start = 0, direction = 1)
```

Here's what we get when we ask for a 2 color palette.

```{r}
show_col(ggc(2))
```

and here's a 6 color palette.

```{r}
show_col(ggc(6))
```

Reproducing the previous plot now with base plot using a grammar of graphics logic:

```{r}
with(data=ant,
{
    # Set margins to achieve a similar aspect-ratio to ggplot
    par(mar=c(4, 4, 0, 4.5) + 0.1)
  
    # Set up coordinate system with default x and y scales and a grid
    plot(x=latitude, y=richness, type="n")
    grid(lty=1, lwd=0.5, col="gray90")
  
    # Add point layer with mappings to visual channels
    points(x=latitude, y=richness, col=ggc(2)[habitat], pch=20)
  
    # Add a legend to the outer margin (by first adding a blank panel)
    par(mar=c(4, 4, 0, 0) + 0.1, new=TRUE)
    plot(1, 1, type="n", axes=FALSE, ann=FALSE)
    legend("right", levels(habitat), col=ggc(2), pch=20, title="habitat", bty="n")
}
)
```

This is essentially the same as the previous plot made by `ggplot`. Clearly base plot is more work than `ggplot` but base plot is good, if sometimes ad hoc, for those times that you need complete control of the graphical elements.


## Training

Now, let's move on to analysis. Fit the full model using `glm()`, which uses a maximum likelihood algorithm:

```{r}
fitHxL <- glm(richness ~ habitat + latitude + habitat:latitude, family=poisson, data=ant)
coef(fitHxL) #Maximum likelihood estimates of parameters
```

Diagnostic plots

```{r}
par(mfrow=c(2,2))
plot(fitHxL,1:4, ask=FALSE)
```

Overall, the diagnostics seem not too bad. It is certainly an improvement over the Normal linear model we tried before. There is perhaps still a slight tendency for the standardized residuals to increase with the mean (we'll investigate that further later) and there is an influential point (case 25). We should investigate the influential point using our case deletion algorithm to examine parameter sensitivity to this point. We'll return to that later as we continue to improve the model.


## Inference

Out of the box inferences about parameter uncertainty come from the assumption of a Normal sampling distribution for the parameter z-values (estimate divided by standard error), giving us a p-value for each parameter.

```{r}
summary(fitHxL)
```
```{r eval=FALSE}
coef(summary(fitHxL))[,1:2]   #or use this for just the coefficients
```

However, most of the parameter hypotheses being considered here are not meaningful, either because we don't care about them biologically (e.g. we don't care about the `intercept` hypothesis that species richness is different from zero at the equator), or because the parameters are all interdependent (e.g. the fact that the `habitatforest` parameter is not significantly different from zero means nothing because it is balanced against the estimate for `habitatforest:latitude`). The one slightly meaningful inference is for the `habitatforest:latitude` parameter, which is telling us that there is little evidence for an interaction on the scale of
the linear predictor. But even this doesn't necessarily translate to the biological hypothesis of no interaction between habitat and latitude since the relationship with latitude is nonlinear, as we will see more clearly when the fitted model is plotted below. Take away: most of the hypothesis tests in these summary tables are usually irrelevant to hypotheses that we are scientifically interested in, except in infrequent special cases. As we'll see later, if we want to test specific scientific hypotheses, we'll need to
be deliberate about expressing them as alternative models.

95% confidence intervals for parameters are obtained by profiling the likelihood surface. This is just as we did earlier by varying one parameter while optimizing over the others. Here, the intervals are frequentist and are formed by assuming that the likelihood ratio has a $\chi ^{2}$ sampling distribution. Note this is not the same as the pure likelihood intervals we constructed earlier since here the likelihood ratio is merely a statistic and the inferential basis is the sampling distribution of the statistic.

```{r}
confint(fitHxL)
```

The correlation matrix for the parameters shows how the `habitatforest` and `habitatforest:latitude` parameters are highly interdependent.

```{r}
cov2cor(vcov(fitHxL))
```

Although meaningless for a single model, the log likelihood is useful for comparison with alternative models. Here's how we can extract the log likelihood.

```{r}
logLik(fitHxL)
```

To construct regression intervals, we can use the `predict()` function. For GLMs, there is no ' interval = "confidence" ' option, so we have to construct intervals from the standard errors. This is approximate. More accurate intervals are obtained by parametric bootstrap.

```{r}
newd <- data.frame(latitude = rep(seq(min(ant$latitude), max(ant$latitude), length.out=100),2),
                   habitat = factor(rep(c("bog","forest"),each=100)))
preds <- predict(fitHxL,newdata=newd,se.fit=TRUE)
mnlp <- preds$fit        #mean of the linear predictor
selp <- preds$se.fit     #se of the linear predictor
cillp <- mnlp - 2 * selp #lower of 95% CI for linear predictor
ciulp <- mnlp + 2 * selp #upper
cil <- exp(cillp)        #lower of 95% CI for response scale
ciu <- exp(ciulp)        #upper
mn <- exp(mnlp)          #mean of response scale
preds <- cbind(newd,preds,cil,ciu,mn)
```

In `ggplot` we can plot this in three layers:

* a ribbon geom (a type of polygon) for intervals, using the `alpha` setting to obtain a nice see-through effect
* a line geom for estimated means
* a point geom for the data

We have two dataframes, one with the data and the other with the predictions, so we need to specify the dataframe for each layer. By giving the point layer second in the sequence, the y-axis label is "richness" (if we put the line layer second, the label is "mn") .

```{r}
ggplot() +
    geom_ribbon(data=preds, mapping=aes(x=latitude, ymin=cil, ymax=ciu, fill=habitat), alpha=0.2) +
    geom_point(data=ant, mapping=aes(x=latitude, y=richness, col=habitat)) +
    geom_line(data=preds, mapping=aes(x=latitude, y=mn, col=habitat))
    
```

Finally, here is an illustration of how we can use base plot with similar styling to the `ggplot` defaults. We'll add three variations to improve visual perception:

* colors with semantic meaning (green for forest, orange/brown for bog) that
are also colorblind friendly
* labels added directly to the plot to reduce translational effort
* y-axis also includes zero

```{r}
# Bog & Forest colors
bfc <- c("#d95f02", "#1b9e77")

# Coordinate system with scales (ticks and lines), and grey background
par(mar=c(5, 4, 0, 4) + 0.1)
with(ant, plot(x=latitude, y=richness, ylim=c(0,18), type="n", axes=FALSE, ann=FALSE))
u <- par("usr") #The coordinates of the plot area
rect(u[1], u[3], u[2], u[4], col="grey92", border=NA) #grey background
axis(1, at=42:45, col=NA, col.ticks="black")
axis(2, col=NA, col.ticks="black", las=2)
abline(v=axTicks(1), col="white", lty=1)
abline(h=axTicks(2), col="white", lty=1)

# A polygon layer for intervals
h <- c("bog","forest")
for ( i in 1:2 ) {
    subd <- subset(preds, habitat==h[i])
    x <- subd$latitude
    ymin <- subd$cil
    ymax <- subd$ciu
    polygon(x=c(x, rev(x)), y=c(ymin, rev(ymax)),
            col=adjustcolor(bfc[i], alpha.f=0.2),
            border=NA)
}

# A line layer for means
for ( i in 1:2 ) {
    subd <- subset(preds, habitat==h[i])
    with(subd, lines(x=latitude, y=mn, col=bfc[i], lty=1))
}

# A points layer for the data
with(ant, points(x=latitude, y=richness, col=bfc[habitat], pch=20))

# Annotations
mtext("Latitude (degrees north)", 1, line=2.5)
mtext("Ant species richness", 2, line=2.5)
text(42.9,3.3, "Bog", col=bfc[1])
text(43.85,9.5, "Forest", col=bfc[2])
```

Compared to `ggplot`, it is clear that we have to do extra subsetting and styling work to achieve the same result but base plot is very flexible if somewhat hacky and ad hoc. Since no packages were used, this plot will be highly reproducible over time.